---
title: "project_v3"
output: html_document
date: '2022-12-02'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
library(shiny)


```

#import the original data
```{r}
life_lad = read_excel("/Users/Jacob/Documents/STAT433/Project/Chapter2OnlineData.xls")
fer_rate = read.csv("/Users/Jacob/Documents/STAT433/Project/children-per-woman-UN.csv")
pollution = read.csv("/Users/Jacob/Documents/STAT433/Project/outdoor-pollution-deaths-1990-2017.csv")
income_ineq = read_excel("/Users/Jacob/Documents/STAT433/Project/DataInput_ChartbookOfEconomicInequality.xls")

sus_rate = read.csv("/Users/Jacob/Documents/STAT433/Project/data.csv")


colnames(life_lad)
```

#clean the data by selecting the columns we want
```{r}
life_lad = life_lad %>% 
  select("Country name", "Year", "Life Ladder", "Log GDP per capita") %>% 
  rename(country = "Country name",
         year = "Year",
         log_gdp = "Log GDP per capita")

fer = fer_rate %>% 
  select(-"Code") %>% 
  rename(ferate = "Fertility.rate...Sex..all...Age..all...Variant..estimates",
         country = "Entity",
         year = "Year")

income_ineqq = income_ineq %>% 
  filter(`meaure of inequality` == "Gini Coefficient") %>% 
  drop_na() %>% 
  mutate(g_coef = value/100) %>% 
  select("country", "year", "g_coef")

pollut = pollution %>% 
  mutate(year = as.numeric(Year)) %>% 
  rename(country = "Entity",
         d = "Deaths...Cause..All.causes...Risk..Outdoor.air.pollution...OWID...Sex..Both...Age..All.Ages..Number.") %>% 
  mutate(death = log(d, base = exp(2))) %>% 
  
  select("country", "year", "death") 
  
  
  

```


```{r}

sus = sus_rate %>% 
  filter(X.1 == "Both sexes") %>% 
  rename(susrate_2019 = "Crude.suicide.rates..per.100.000.population.",
         susrate_2018 = "Crude.suicide.rates..per.100.000.population..1" ,
         susrate_2017 = "Crude.suicide.rates..per.100.000.population..2",
         susrate_2016 = "Crude.suicide.rates..per.100.000.population..3" ,
         susrate_2015 = "Crude.suicide.rates..per.100.000.population..4",
         susrate_2014 = "Crude.suicide.rates..per.100.000.population..5",
         susrate_2013 = "Crude.suicide.rates..per.100.000.population..6",
         susrate_2012 = "Crude.suicide.rates..per.100.000.population..7",
         susrate_2011 = "Crude.suicide.rates..per.100.000.population..8",
         susrate_2010 = "Crude.suicide.rates..per.100.000.population..9",
         susrate_2009 = "Crude.suicide.rates..per.100.000.population..10",
         susrate_2008 = "Crude.suicide.rates..per.100.000.population..11",
         susrate_2007 = "Crude.suicide.rates..per.100.000.population..12",
         susrate_2006 = "Crude.suicide.rates..per.100.000.population..13",
         susrate_2005 = "Crude.suicide.rates..per.100.000.population..14",
         susrate_2004 = "Crude.suicide.rates..per.100.000.population..15",
         susrate_2003 = "Crude.suicide.rates..per.100.000.population..16",
         susrate_2002 = "Crude.suicide.rates..per.100.000.population..17",
         susrate_2001 = "Crude.suicide.rates..per.100.000.population..18",
         susrate_2000 = "Crude.suicide.rates..per.100.000.population..19" ,
         country = "X") %>% 
  pivot_longer(cols = c("susrate_2019", "susrate_2018", "susrate_2017", "susrate_2016", "susrate_2015", "susrate_2014", "susrate_2013", "susrate_2012", "susrate_2011", "susrate_2010", "susrate_2009", "susrate_2008", "susrate_2007", "susrate_2006", "susrate_2005", "susrate_2004", "susrate_2003", "susrate_2002", "susrate_2001", "susrate_2000"), names_to = "Year", values_to = "susrate") %>% 
  separate(Year, c('x', 'y')) %>% 
  mutate(year = as.numeric(y)) %>% 
  mutate(r = substr(susrate, 1, 3) ) %>% 
  mutate(rate = as.numeric(r)) %>% 
  select("country", "year", "rate")
  

str(sus)
```

#Join the clean data
```{r}
df = life_lad %>% 
  rename(ll = "Life Ladder") %>% 
  left_join(fer, by = c("country", "year")) %>% 
  left_join(sus, by = c("country", "year")) %>% 
  left_join(income_ineqq, by = c("country", "year")) %>% 
  left_join(pollut, by = c("country", "year"))

```

#Plot the correlation of mean value
```{r}
df1 = df %>% 
  group_by(country) %>% 
  summarise(mean_ll = mean(ll),
            mean_fer = mean(ferate, na.rm = T),
            mean_sucide = mean(rate, na.rm = T),
            mean_d_pollut = mean(death, na.rm = T),
            mean_g = mean(g_coef, na.rm = T))
  
full_df <- gather(df1, key="measure", value="value", c("mean_fer", "mean_sucide", "mean_d_pollut", "mean_g"))



ggplot(full_df, aes(x=mean_ll, y=value))+
  geom_point()+
  facet_wrap(~measure, ncol = 2, scales = "free_y")
  
  

```

#fit the linear model
```{r}
life_lm_f = lm(mean_ll ~ mean_fer, data=df1)
life_lm_p = lm(mean_ll ~ mean_d_pollut, data=df1)
life_lm_s = lm(mean_ll ~ mean_sucide, data=df1)
life_lm_g = lm(mean_ll ~ mean_g, data=df1)

summary(life_lm_f)
summary(life_lm_p)
summary(life_lm_s)
summary(life_lm_g)






```

#outlier of susicide rate
```{r}
df1 %>% 
  arrange(desc(mean_sucide))

#The country is Lesotho

```

# Geographic distribution of countries that has fertilty rate above 4
```{r}
df_f4 = df1 %>% 
  filter(mean_fer >=4) %>% 
  select("country", "mean_fer") %>% 
  rename("value" = "mean_fer")

world2 <- map_data("world")

world2 %>%
  merge(df_f4, by.x = "region", by.y = "country", all.x = T) %>%
  arrange(group, order) %>%
  ggplot(aes(x = long, y = lat, group = group, fill = value)) + geom_polygon()


```


#Geographic dist of life ladder
```{r}
ddf = df1 %>% 
  select("country", "mean_ll") %>% 
  rename("value" = "mean_ll")

world <- map_data("world")

world %>%
  merge(ddf, by.x = "region", by.y = "country", all.x = T) %>%
  arrange(group, order) %>%
  ggplot(aes(x = long, y = lat, group = group, fill = value)) + geom_polygon()



```

#Geographic dist of fertility rate
```{r}
ddf1 = df1 %>% 
  select("country", "mean_fer") %>% 
  rename("value" = "mean_fer")

world1 <- map_data("world")

world1 %>%
  merge(ddf1, by.x = "region", by.y = "country", all.x = T) %>%
  arrange(group, order) %>%
  ggplot(aes(x = long, y = lat, group = group, fill = value)) + geom_polygon()



```

#Controlling GDP (lower)
```{r}
max_gdp = max(df$log_gdp, na.rm = T)
min_gdp = min(df$log_gdp, na.rm = T)
mean_gdp = mean(df$log_gdp, na.rm = T)
  
df_lower_gdp = df %>% 
  filter(log_gdp >= min_gdp & log_gdp <= mean_gdp) %>% 
  group_by(country) %>% 
  summarise(mean_ll = mean(ll),
            mean_fer = mean(ferate, na.rm = T),
            mean_sucide = mean(rate, na.rm = T),
            mean_d_pollut = mean(death, na.rm = T),
            mean_g = mean(g_coef, na.rm = T))
  
full_df_lgdp <- gather(df_lower_gdp, key="measure", value="value", c("mean_fer", "mean_sucide", "mean_d_pollut", "mean_g"))



ggplot(full_df_lgdp, aes(x=mean_ll, y=value))+
  geom_point()+
  facet_wrap(~measure, ncol = 2, scales = "free_y")

life_lm_fl = lm(mean_ll ~ mean_fer, data=df_lower_gdp)
life_lm_pl = lm(mean_ll ~ mean_d_pollut, data=df_lower_gdp)
life_lm_sl = lm(mean_ll ~ mean_sucide, data=df_lower_gdp)
life_lm_gl = lm(mean_ll ~ mean_g, data=df_lower_gdp)

summary(life_lm_fl)
summary(life_lm_pl)
summary(life_lm_sl)
summary(life_lm_gl)

#For lower gdp country, negative correlation between life ladder and fertility rate is still significant
```

#Controlling GDP (higher)
```{r}
df_higher_gdp = df %>% 
  filter(log_gdp >= mean_gdp & log_gdp <= max_gdp) %>% 
  group_by(country) %>% 
  summarise(mean_ll = mean(ll),
            mean_fer = mean(ferate, na.rm = T),
            mean_sucide = mean(rate, na.rm = T),
            mean_d_pollut = mean(death, na.rm = T),
            mean_g = mean(g_coef, na.rm = T))
  
full_df_hgdp <- gather(df_higher_gdp, key="measure", value="value", c("mean_fer", "mean_sucide", "mean_d_pollut", "mean_g"))



ggplot(full_df_hgdp, aes(x=mean_ll, y=value))+
  geom_point()+
  facet_wrap(~measure, ncol = 2, scales = "free_y")

life_lm_fh = lm(mean_ll ~ mean_fer, data=df_higher_gdp)
life_lm_ph = lm(mean_ll ~ mean_d_pollut, data=df_higher_gdp)
life_lm_sh = lm(mean_ll ~ mean_sucide, data=df_higher_gdp)
life_lm_gh = lm(mean_ll ~ mean_g, data=df_higher_gdp)

summary(life_lm_fh)
summary(life_lm_ph)
summary(life_lm_sh)
summary(life_lm_gh)

# In higher gdp countries, poolution, fertility rate and income inequality all have significant correlation between life ladder


```



#Find out the year range of each country
```{r}
df_year = df %>% 
  group_by(country) %>% 
  summarise(max_year = max(year),
            min_year = min(year),
            year_range = max_year - min_year)
 
df_min = df %>% 
  left_join(df_year) %>% 
  filter(year_range >=10,
         year == min_year) %>% 
  rename(ll_min = ll,
         fer_min = ferate,
         sus_min = rate,
         pollut_min = death,
         g_min = g_coef) %>% 
  select(-"year_range", -"max_year", -"min_year", -"year", "log_gdp")
  
  
df_diff = df %>% 
  left_join(df_year) %>% 
  filter(year_range >=10,
         year == min_year + 10) %>% 
  rename(ll_max = ll,
         fer_max = ferate,
         sus_max = rate,
         pollut_max = death,
         g_max = g_coef) %>% 
  select(-"year_range", -"max_year", -"min_year", -"year", "log_gdp") %>% 
  left_join(df_min, by = "country") %>% 
  mutate(ll_diff = ll_max - ll_min,
         fer_diff = fer_max - fer_min,
         sus_diff = sus_max - sus_min,
         pollut_diff = pollut_max - pollut_min,
         g_diff = g_max - g_min) %>% 
  select("country", "ll_diff", "fer_diff", "sus_diff", "pollut_diff", "g_diff") 
  
  

full_df_diff <- gather(df_diff, key="measure", value="value", c("fer_diff", "sus_diff", "pollut_diff", "g_diff"))



ggplot(full_df_diff, aes(x=ll_diff, y=value))+
  geom_point()+
  facet_wrap(~measure, ncol = 2, scales = "free_y")




```


#fit the model
```{r}
life_lm_time_f = lm(ll_diff ~ fer_diff, data=df_diff)
life_lm_time_s = lm(ll_diff ~ sus_diff, data=df_diff)
life_lm_time_p = lm(ll_diff ~ pollut_diff, data=df_diff)


summary(life_lm_time_f)
summary(life_lm_time_s)
summary(life_lm_time_p)





#Not significant
```

#We are expected some drastic changes in the plot, but we could see that most lines are smooth
```{r}

df_seq_ll = df %>% 
   left_join(df_year) %>% 
  filter(year_range >=10,
         year >= min_year & year <= min_year + 10) %>% 
  select("country", "year", "ll", "ferate") %>% 
  mutate(ratio = ll/ferate)
  
  

ggplot(df_seq_ll, aes(x = year, y = ratio, col = country))+
  geom_smooth()+
  theme(legend.position = "none")




```

```{r}
df_seq_s = df %>% 
   left_join(df_year) %>% 
  filter(year_range >=10,
         year >= min_year & year <= min_year + 10) %>% 
  select("country", "year", "ll", "rate") %>% 
  mutate(ratio = ll/rate)


ggplot(df_seq_s, aes(x = year, y = ratio, col = country))+
  geom_smooth()+
  theme(legend.position = "none")
```

```{r}
#runApp("/Users/Jacob/Documents/STAT433/Project/app.R")
runGitHub("STAT433-F22", "jwhitedubs")
```